---
title: "Big Data Cup Women's Hockey"
author: "Ben Howell"
date: "1/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE,
                      fig.align = "center")
```

Event %in% c("Faceoff Win", "Takeaway", "Puck Recovery") for events that start a possession w/ row_number
1:18:00 moodboards stream: https://www.twitch.tv/videos/902736929

```{r Load/Manipulate Data}
library(tidyverse)
library(grid)
library(gganimate)
library(gifski)
library(mgcv)
library(caret)
library(nnet)
library(Matrix)
library(Ckmeans.1d.dp)
library(mclust)
library(broom)
library(ggplot2)
library(png)
library(grid)
library(raster)
library(magick)
library(mlr)
library(knitr)
library(cmna)
library(xgboost)

bdc <- read.csv("https://raw.githubusercontent.com/bigdatacup/Big-Data-Cup-2021/main/hackathon_womens.csv") %>%
  mutate(League = "BDC")
nwhl <- read.csv("https://raw.githubusercontent.com/bigdatacup/Big-Data-Cup-2021/main/hackathon_nwhl.csv") %>%
  mutate(League = "NWHL")

data <- rbind(bdc, nwhl)

table(data$Event)
goal_x <- c(188, 190, 190, 188)
goal_y <- c(40, 40, 45, 45)

goal <- data.frame(goal_x, goal_y)

rink <- function(...) {
  ggplot(...)  +
  geom_segment(aes(x = 25 - 100, xend = 175 - 100, y = 0 - 42.5, yend = 0 - 42.5)) +
  geom_segment(aes(x = 25 - 100, xend = 175 - 100, y = 85 - 42.5, yend = 85 - 42.5)) +
  geom_segment(aes(x = 0 - 100, xend = 0 - 100, y = 20 - 42.5, yend = 60 - 42.5)) +
  geom_segment(aes(x = 200 - 100, xend = 200 - 100, y = 20 - 42.5, yend = 60 - 42.5)) +
  geom_curve(aes(x = 0 - 100, xend = 25 - 100, y = 60 - 42.5, yend = 85 - 42.5),
             curvature = -0.32) +
  geom_curve(aes(x = 0 - 100, xend = 25 - 100, y = 20 - 42.5, yend = 0 - 42.5),
             curvature = 0.32) +
  geom_curve(aes(x = 175 - 100, xend = 200 - 100, y = 0 - 42.5, yend = 20 - 42.5),
             curvature = 0.32) +
  geom_curve(aes(x = 175 - 100, xend = 200 -100, y = 85 - 42.5, yend = 60 - 42.5),
             curvature = -0.32)  +
  geom_path(data = goal, aes(x = goal_x - 100, y = goal_y - 42.5)) +
  geom_segment(aes(x = 100 - 100, xend = 100 - 100, y = 85 - 42.5, yend = 0 - 42.5), size = 2, color = "red")
}

num <- nrow(data)

Minute <- 0:20
min_of_game <- 20:0

time <- data.frame(Minute, min_of_game)

data <- data %>%
  mutate(X.Coordinate = X.Coordinate - 100,
         X.Coordinate.2 = X.Coordinate.2 - 100,
         Y.Coordinate = Y.Coordinate - 42.5,
         Y.Coordinate.2 = Y.Coordinate.2 - 42.5,
         row_num = row_number(),
         Start_end = ifelse(Period == "1" & Clock == "20:00" & Event == "Faceoff Win", "Start", ""), #creates start of game and end of game variables
         Start_end = ifelse((lead(Start_end) == "Start" & Period >= 3) | (row_num >= num), "End", 
                            paste(Start_end)),
         change_poss_two = ifelse(Start_end != "", paste(Start_end), 
                              ifelse(lead(Team) != Team, "Turnover", "No Change")),
         change_poss = ifelse(Start_end != "", "No Change", 
                              ifelse(lead(Team) != Team, "Turnover", "No Change")),
         Clock_two = paste(as.character(Clock))) %>%
  separate(Clock_two, c("Minute", "Second"), sep = ":") %>%
  mutate(Minute = as.numeric(as.character(Minute)))

data <- right_join(data, time, by = "Minute")

data <- data %>%
  mutate(min_of_play = ifelse(Period == 2, min_of_game + 20,
                              ifelse(Period == 3, min_of_game + 40,
                                     ifelse(Period == 4, min_of_game + 60,
                                            min_of_game + 0))),
         Total_goals = Home.Team.Goals + Away.Team.Goals,
         Event_tm_skaters = ifelse(Team == Home.Team, Home.Team.Skaters,
                                   Away.Team.Skaters),
         Opponent_tm_skaters = ifelse(Team == Home.Team, Away.Team.Skaters, Home.Team.Skaters))

end <- data %>%
  filter(Start_end %in% c("End")) %>%
  dplyr::select(game_date, Home.Team, Away.Team, Home.Team.Goals, Away.Team.Goals) %>%
  mutate(Winner = ifelse(Home.Team.Goals > Away.Team.Goals, paste(Home.Team),
                         ifelse(Home.Team.Goals < Away.Team.Goals, paste(Away.Team),
                                ifelse(Home.Team.Goals == Away.Team.Goals, "Tie", "Ugh"))),
         Winner = ifelse(game_date == "2018-10-19", paste(Home.Team), paste(Winner))) %>%
  dplyr::select(-Home.Team.Goals, -Away.Team.Goals)
#manually set St. Lawrence as winner on 2018-10-19
#game went to sudden overtime and final event recorded (a goal) is at a 3-3 tie

data <- right_join(data, end, by = c("game_date", "Home.Team", "Away.Team"))

store <- data 

store <- store %>%
  filter(Event %in% c("Faceoff Win", "Takeaway", "Puck Recovery")) %>%
  mutate(Play_event = row_number()) #begin of a sequence of events for a team
#ignores the potential for dump and chase w/ a Puck Recovery

data <- left_join(data, store)

data <- data %>%
  fill(Play_event) #put sequences of plays together

data <- data %>%
  mutate(Goal_diff = Home.Team.Goals - Away.Team.Goals,
         Win = ifelse(Winner == Home.Team, "Yes", "No"),
         Event_tm_skaters = ifelse(Team == Home.Team, Home.Team.Skaters,
                                   Away.Team.Skaters),
         Opponent_tm_skaters = ifelse(Team == Home.Team, Away.Team.Skaters, Home.Team.Skaters),
         Skate_diff = Event_tm_skaters - Opponent_tm_skaters,
         label = ifelse(Event == "Goal", 1, 0),
         label = as.factor(label))
```

# xgBoost xG Model

```{r xgboost xG Model}
# walkthrough and tutorial: https://www.datatechnotes.com/2020/08/regression-example-with-xgboost-in-r.html
train_games <- data %>%
  filter(League %in% c("NWHL")) 
#split into testing and training games so that we can keep passing sequences intact
test_games <- data %>%
  filter(League == "BDC")

train_shots <- train_games %>%
  filter(Event %in% c("Shot", "Goal") & Detail.2 != "Blocked") %>%
  dplyr::select(Event_tm_skaters, Opponent_tm_skaters, min_of_play, X.Coordinate, Y.Coordinate, 
                Detail.1, Detail.2, Detail.3, Detail.4, Skate_diff, label)

test_shots <- test_games %>%
  filter(Event %in% c("Shot", "Goal") & Detail.2 != "Blocked") %>%
  dplyr::select(Event_tm_skaters, Opponent_tm_skaters, min_of_play, X.Coordinate, Y.Coordinate, 
                Detail.1, Detail.2, Detail.3, Detail.4, Skate_diff, label) 
#removed blocked shots from the dataset

train_x <- data.matrix(train_shots[, -11])
train_y <- train_shots[, 11]

test_x <- data.matrix(test_shots[, -11])
test_y <- test_shots[, 11]

xgb_train <- xgb.DMatrix(data = train_x, label = as.numeric(as.character(train_y)))
xgb_test <- xgb.DMatrix(data = test_x, label = as.numeric(as.character(test_y)))

xgb_train
xgb_test
```

```{r}
params <- list(booster = "gbtree", objective = "binary:logistic", eta = 0.3,
               max_depth = 8)
set.seed(25)
xgbcv <- xgb.cv(params = params, data = xgb_train, nrounds = 100, nfold = 5, showsd = T, stratified = T, 
                print_every_n = 10, early_stopping_rounds = 20, maximize = F)
#need to figure out what the 'metrics' call here is doing, b/c it changes the output
#will want to try "error"
##best iteration = 10
xgbcv
set.seed(2)
xgbc <- xgb.train(data = xgb_train, params = params, nrounds = 13)
summary(xgbc)
print(xgbc)

pred_y <- predict(xgbc, newdata = xgb_test)

mse <- mean((as.numeric(as.character(test_y)) - pred_y) ^ 2)
mae <- MAE(as.numeric(as.character(test_y)), pred_y)
rmse <- RMSE(as.numeric(as.character(test_y)), pred_y)

x = 1:length(as.numeric(as.character(test_y)))
{plot(x, as.numeric(as.character(test_y)), col = "red", type = "l")
lines(x, pred_y, col = "blue", type = "l")
legend(x = 1, y = 38,  legend = c("original test_y", "predicted test_y"), 
       col = c("red", "blue"), box.lty = 1, cex = 0.8, lty = c(1, 1))}

quantile(pred_y, c(0.25, .5, 0.75, 0.85, 0.9, 0.95))
xgbpred_factors <- ifelse(pred_y > 0.25, 1, 0)
xgbpred_factors <- as.factor(xgbpred_factors)

confusionMatrix(xgbpred_factors, test_y) #gets almost every single missed goal correct, but struggles with predicting goals?
#######################
test_x_df <- as.data.frame(test_x)
preds <- as.data.frame(pred_y)
test_x_df$pred <- preds
test_x_df <- test_x_df %>%
  mutate(xG = (as.numeric(unlist(preds)))) %>%
  dplyr::select(Event_tm_skaters:Skate_diff, xG)
```

```{r}
mid_point <- mean(test_x_df$xG)

test_x_df %>%
  rink() +
  geom_point(aes(x = X.Coordinate, y = Y.Coordinate, color = xG), size = 3) +
  scale_color_gradient(low = "white", high = "red") +
  labs(caption = "Ben Howell | @benhowell71 | benhowell71.com") + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))
  # scale_color_gradient2(low = "blue", mid = "beige", midpoint = 0.2, high = "red") +
  # theme_minimal()

test_x_df %>%
  rink() +
  stat_summary_hex(aes(x = X.Coordinate, y = Y.Coordinate, z = xG),
                   fun = mean, binwidth = c(5, 5)) +
  scale_fill_gradient2(low = "lightblue", mid = "white", midpoint = 0.1, high = "red") +
  theme_minimal() +
  labs(title = "Average xG by Shot Location",
       caption = "Ben Howell | @benhowell71 | benhowell71.com",
       subtitle = "Trained on NWHL data. Tested on Olympic data",
       fill = "Avg xG") +
  geom_path(data = goal, aes(x = goal_x - 100, y = goal_y - 42.5), size = 2) +
  theme(panel.grid = element_blank(), 
        axis.text = element_blank(),
        axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
        plot.caption = element_text(hjust = 0.5, face = "italic", size = 10),
        plot.subtitle = element_text(hjust = 0.5))

import <- xgb.importance(model = xgbc)
xgb.ggplot.importance(importance_matrix = import) +
  theme_minimal() +
  labs(caption = "Ben Howell | @benhowell71 | benhowell71.com") + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))
#Detail 1 = shottype
#Detail 2 = location (on net or missed)
#Traffic
#one timer
```

```{r}
train <- train_games %>%
  filter(! Event %in% c("Shot", "Goal") | Detail.2 == "Blocked") %>%
  mutate(xG = 0)

test <- test_games %>%
  filter(! Event %in% c("Shot", "Goal") | Detail.2 == "Blocked") %>% #rest of events in tran/test games
  mutate(xG = 0)
  
train_shots_context <- train_games %>%
  filter(Event %in% c("Shot", "Goal") & Detail.2 != "Blocked") %>%
  dplyr::select(-Event_tm_skaters, -Opponent_tm_skaters, -min_of_play, -X.Coordinate, -Y.Coordinate, 
                -Detail.1, -Detail.2, -Detail.3, -Detail.4, -Skate_diff, -label)

test_shots_context <- test_games %>%
  filter(Event %in% c("Shot", "Goal") & Detail.2 != "Blocked") %>%
  dplyr::select(-Event_tm_skaters, -Opponent_tm_skaters, -min_of_play, -X.Coordinate, -Y.Coordinate, 
                -Detail.1, -Detail.2, -Detail.3, -Detail.4, -Skate_diff, label)

level_1 <- levels(data$Detail.1)
level_2 <- levels(data$Detail.2)
level_3 <- levels(data$Detail.3)
level_4 <- levels(data$Detail.4)

det_one <- as.data.frame(level_1) %>%
  mutate(One_factor = row_number(),
         Detail.1 = paste(level_1)) %>%
  dplyr::select(-level_1)

det_two <- as.data.frame(level_2) %>%
  mutate(Two_factor = row_number(),
         Detail.2 = paste(level_2)) %>%
  dplyr::select(-level_2)

det_three <- as.data.frame(level_3) %>%
  mutate(Three_factor = row_number(),
         Detail.3 = paste(level_3)) %>%
  dplyr::select(-level_3)

det_four <- as.data.frame(level_4) %>%
  mutate(Four_factor = row_number(),
         Detail.4 = paste(level_4)) %>%
  dplyr::select(-level_4)

test_shots_context$Event_tm_skaters <- test_x_df$Event_tm_skaters
test_shots_context$Opponent_tm_skaters <- test_x_df$Opponent_tm_skaters
test_shots_context$min_of_play <- test_x_df$min_of_play
test_shots_context$X.Coordinate <- test_x_df$X.Coordinate
test_shots_context$Y.Coordinate <- test_x_df$Y.Coordinate
test_shots_context$Detail.1_fac <- test_x_df$Detail.1
test_shots_context$Detail.2_fac <- test_x_df$Detail.2
test_shots_context$Detail.3_fac <- test_x_df$Detail.3
test_shots_context$Detail.4_fac <- test_x_df$Detail.4
test_shots_context$Skate_diff <- test_x_df$Skate_diff
test_shots_context$xG <- test_x_df$xG

test_shots_context <- right_join(test_shots_context, det_one, by = c("Detail.1_fac" = "One_factor"))
test_shots_context <- right_join(test_shots_context, det_two, by = c("Detail.2_fac" = "Two_factor"))
test_shots_context <- right_join(test_shots_context, det_three, by = c("Detail.3_fac" = "Three_factor"))
test_shots_context <- right_join(test_shots_context, det_four, by = c("Detail.4_fac" = "Four_factor"))

test_shots_context <- test_shots_context %>%
  dplyr::select(-Detail.1_fac, -Detail.2_fac, -Detail.3_fac, -Detail.4_fac)

test_df <- rbind(test, test_shots_context) %>%
  filter(!is.na(game_date))

test_df <- test_df %>%
  arrange(Play_event)
```


```{r}
#running xG back on the training dataset
set.seed(22)
pred_x <- predict(xgbc, newdata = xgb_train)

train_x_df <- as.data.frame(train_x)
preds_x <- as.data.frame(pred_x)
train_x_df$pred <- preds_x
train_x_df <- train_x_df %>%
  mutate(xG = (as.numeric(unlist(preds_x)))) %>%
  dplyr::select(Event_tm_skaters:Skate_diff, xG)

train_shots_context <- train_shots_context %>%
  filter(! is.na(game_date))

train_shots_context$Event_tm_skaters <- train_x_df$Event_tm_skaters
train_shots_context$Opponent_tm_skaters <- train_x_df$Opponent_tm_skaters
train_shots_context$min_of_play <- train_x_df$min_of_play
train_shots_context$X.Coordinate <- train_x_df$X.Coordinate
train_shots_context$Y.Coordinate <- train_x_df$Y.Coordinate
train_shots_context$Detail.1_fac <- train_x_df$Detail.1
train_shots_context$Detail.2_fac <- train_x_df$Detail.2
train_shots_context$Detail.3_fac <- train_x_df$Detail.3
train_shots_context$Detail.4_fac <- train_x_df$Detail.4
train_shots_context$Skate_diff <- train_x_df$Skate_diff
train_shots_context$xG <- train_x_df$xG

train_shots_context <- right_join(train_shots_context, det_one, by = c("Detail.1_fac" = "One_factor"))
train_shots_context <- right_join(train_shots_context, det_two, by = c("Detail.2_fac" = "Two_factor"))
train_shots_context <- right_join(train_shots_context, det_three, by = c("Detail.3_fac" = "Three_factor"))
train_shots_context <- right_join(train_shots_context, det_four, by = c("Detail.4_fac" = "Four_factor"))

train_shots_context <- train_shots_context %>%
  dplyr::select(-Detail.1_fac, -Detail.2_fac, -Detail.3_fac, -Detail.4_fac)

train_shots_context <- train_shots_context %>%
  filter(! is.na(game_date)) %>%
  mutate(label = ifelse(Event == "Goal", 1, 0))

train_df <- rbind(train, train_shots_context) %>%
  filter(!is.na(game_date))

train_df <- train_df %>%
  arrange(Play_event)

test_df <- rbind(test_df, train_df)
```

as reference: https://karun.in/blog/expected-threat.html
- divide the rink into zones
  + essentially a 200 x 85 rink
- start with worth 0 for all zones and pass types, then evaluate formula **iteratively until convergence**
  + can then evaluate the formula based on xV from previous iteration

# Rink Stuff

```{r}
######
xbreaks <- seq(-100.01, 100.01, length.out = 11)
#goes from -100.01 to 100.01 to properly account values at 100 or -100
ybreaks <- seq(-42.6, 42.6, length.out = 8)
#goes from -42.6 to 42.6 to properly account values at 42.5 or -42.5

id <- matrix(seq_len(length(xbreaks) * length(ybreaks)),
             length(xbreaks), length(ybreaks))

xi <- findInterval(data$X.Coordinate, xbreaks)
yi <- findInterval(data$Y.Coordinate, ybreaks)

data$id <- id[cbind(xi, yi)]

x2 <- findInterval(data$X.Coordinate.2, xbreaks)
y2 <- findInterval(data$Y.Coordinate.2, ybreaks)

data$id_two <- id[cbind(x2, y2)]

###########needed for later as graphing references
xbrk <- seq(-90, 90, length.out = 10)
ybrk <- seq(-36.43, 36.43, length.out = 7)

points <- merge(xbrk, ybrk)

x_p <- findInterval(points$x, xbreaks)
y_p <- findInterval(points$y, ybreaks)

points$id <- id[cbind(x_p, y_p)]

points <- points %>%
  mutate(x_min = x - 10, 
         x_max = x + 10, 
         y_min = y - 6.07,
         y_max = y + 6.07)

data <- data %>%
  ungroup() %>%
  group_by(Play_event) %>%
  mutate(next_zone = lead(id),
         zone_next = ifelse(is.na(next_zone), id_two, next_zone))
  #so, for unsuccessful passes, we take the intended location or zone
  #allows measuring how successful passes b/t zones are
  #though, when we measure for xAV, unsuccessful passes receive a value of 0
  #but that's a bridge to cross when we get there
#
```

```{r}
#Zone, AVzone, Movement%
zone <- data %>%
  filter(Event %in% c("Goal", "Shot", "Play", "Incomplete Play")) %>%
  #not considering dump in/outs b/c intentionally conceding possession
  group_by(id) %>%
  count(Event) %>%
  spread(Event, n, fill = 0) %>%
  mutate(Events = Goal + `Incomplete Play` + Shot + Play,
         Shots = Goal + Shot,
         #creating separate shots variable makes shooting% calculation easier
         Shot_pct = round(Shots / Events, digits = 3),
         Move_pct = round((`Incomplete Play` + Play) / Events, digits = 3),
         #Shot_pct + Move_pct should equal 1
         Make_pct = round(Goal / Shots, digits = 3),
         Make_pct = ifelse(Make_pct == "NaN", 0, Make_pct),
         #cleaning up the Make_pct variable so that we have 0s instead of NaNs
         AVxy = round(Shot_pct * Make_pct, digits = 5)) %>%
  dplyr::select(id, Shot_pct, Make_pct, Move_pct)
```

```{r}
#movement between zones
movement <- data %>%
  filter(Event %in% c("Goal", "Shot", "Play", "Incomplete Play")) %>%
  #not considering dump in/outs b/c intentionally conceding possession
  group_by(id) %>%
  count(zone_next) %>%
  #using zone_next b/c it accounts for turnovers
  #since we want to know how often someone tries to pass from Zone X to Zone Y
  #knowing where they were aimed helps, even if the pass was taken away
  filter(! is.na(zone_next))

obs <- data %>%
  filter(! is.na(zone_next)) %>%
  ungroup() %>%
  count(id) %>%
  mutate(Obs = n) %>%
  dplyr::select(-n)

movement <- right_join(movement, obs, by = c("id"))
#enables us to find out how often someone went from Zone id to the next zone
#essentially: movement / total movements from that zone

movement <- movement %>%
  mutate(pct_of_moves = round(n / Obs, digits = 5)) %>%
  dplyr::select(id, zone_next, pct_of_moves)
```

```{r}
success <- data %>%
  filter(Event %in% c("Goal", "Shot", "Play", "Incomplete Play")) %>%
  #not considering dump in/outs b/c intentionally conceding possession
  group_by(id, zone_next) %>%
  count(change_poss) %>%
  filter(! is.na(zone_next)) %>%
  spread(change_poss, n, fill = 0) %>%
  mutate(Passes = `No Change` + Turnover,
         Success_pct = round(`No Change` / Passes, digits = 5)) %>%
  dplyr::select(id, zone_next, Success_pct)

nrow(movement) == nrow(success)
#should be true
```

```{r}
movement <- right_join(movement, success, by = c("id", "zone_next"))

all <- right_join(movement, zone, by = c("id"))

zw <- zone %>%
  #this will give us the AVzw for iteration two
  #iteration 1 was a simple shot * make
  rename(c("zone_next" = "id",
           "zw_shot_pct" = "Shot_pct",
           "zw_make_pct" = "Make_pct")) %>%
  dplyr::select(-Move_pct)

all <- right_join(all, zw, by = c("zone_next"))
```

```{r}
it_two <- all %>%
  mutate(AVzw = zw_shot_pct * zw_make_pct,
         move_score = Success_pct * pct_of_moves,
         value = move_score * AVzw)

score <- it_two %>%
  group_by(id) %>%
  #sum of scores from zone that you are in
  summarise(SUM_VAL = sum(value))

total <- right_join(zone, score, by = c("id"))

total <- total %>%
  mutate(xAVxy_one = round((Shot_pct * Make_pct) +
                      (Move_pct * SUM_VAL), digits = 4))
```

### Next Iteration

```{r}
constants <- it_two %>%
  dplyr::select(id, zone_next, Shot_pct, Make_pct, Move_pct, move_score)
#need to add in xAVxy from above as xAVzw
#i.e. evaluate in iterations
one_res <- total %>%
  dplyr::select(id, xAVxy_one) %>%
  rename(c("AVzw" = "xAVxy_one",
           "zone_next" = "id"))

it_three <- right_join(constants, one_res, by = c("zone_next"))

it_three <- it_three %>%
  mutate(value = move_score * AVzw)

score_three <- it_three %>%
  group_by(id) %>%
  summarise(SUM_VAL = sum(value)) %>%
  right_join(zone, by = c("id")) %>%
  mutate(AV_two = round((Shot_pct * Make_pct) +
                      (Move_pct * SUM_VAL), digits = 4))
```

### Fourth Iteration

```{r}
three_res <- score_three %>%
  dplyr::select(id, AV_two) %>%
  rename(c("AVzw" = "AV_two",
           "zone_next" = "id"))

it_four <- right_join(constants, three_res, by = c("zone_next"))

it_four <- it_four %>%
  mutate(value = move_score * AVzw)

score_four <- it_four %>%
  group_by(id) %>%
  summarise(SUM_VAL = sum(value)) %>%
  right_join(zone, by = c("id")) %>%
  mutate(AV_four = round((Shot_pct * Make_pct) +
                      (Move_pct * SUM_VAL), digits = 4))
```

### Fifth Iteration

```{r}
four_res <- score_four %>%
  dplyr::select(id, AV_four) %>%
  rename(c("AVzw" = "AV_four",
           "zone_next" = "id"))

it_five <- right_join(constants, four_res, by = c("zone_next"))

it_five <- it_five %>%
  mutate(value = move_score * AVzw)

score_five <- it_five %>%
  group_by(id) %>%
  summarise(SUM_VAL = sum(value)) %>%
  right_join(zone, by = c("id")) %>%
  mutate(AV_five = round((Shot_pct * Make_pct) +
                      (Move_pct * SUM_VAL), digits = 4))
```

```{r}
results <- score_five %>%
  dplyr::select(id, AV_five) %>%
  rename(c("xAV" = "AV_five")) %>%
  right_join(points, by = c("id"))
#wow! Zones right in front of the goal have the highest xAV
```

```{r}
med <- median(results$xAV)

results %>%
  ggplot() +
  geom_rect(aes(xmin = x_min, xmax = x_max, ymin = y_min, ymax = y_max, color = xAV, fill = xAV)) +
  scale_color_gradient2(low = "lightblue", mid = "white", midpoint = med, high = "red") +
  scale_fill_gradient2(low = "lightblue", mid = "white", midpoint = med, high = "red")  +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank())  +
  geom_segment(aes(x = 25 - 100, xend = 175 - 100, y = 0 - 42.5, yend = 0 - 42.5)) +
  geom_segment(aes(x = 25 - 100, xend = 175 - 100, y = 85 - 42.5, yend = 85 - 42.5)) +
  geom_segment(aes(x = 0 - 100, xend = 0 - 100, y = 20 - 42.5, yend = 60 - 42.5)) +
  geom_segment(aes(x = 200 - 100, xend = 200 - 100, y = 20 - 42.5, yend = 60 - 42.5)) +
  geom_curve(aes(x = 0 - 100, xend = 25 - 100, y = 60 - 42.5, yend = 85 - 42.5),
             curvature = -0.32) +
  geom_curve(aes(x = 0 - 100, xend = 25 - 100, y = 20 - 42.5, yend = 0 - 42.5),
             curvature = 0.32) +
  geom_curve(aes(x = 175 - 100, xend = 200 - 100, y = 0 - 42.5, yend = 20 - 42.5),
             curvature = 0.32) +
  geom_curve(aes(x = 175 - 100, xend = 200 -100, y = 85 - 42.5, yend = 60 - 42.5),
             curvature = -0.32)  +
  geom_path(data = goal, aes(x = goal_x - 100, y = goal_y - 42.5)) +
  geom_segment(aes(x = 100 - 100, xend = 100 - 100, y = 85 - 42.5, yend = 0 - 42.5), size = 2, color = "red")  +
  labs(title = "Average xAV by Location",
       caption = "Ben Howell | @benhowell71 | benhowell71.com",
       fill = "xAV") +
  geom_path(data = goal, aes(x = goal_x - 100, y = goal_y - 42.5), size = 2) +
  theme(panel.grid = element_blank(), 
        axis.text = element_blank(),
        axis.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
        plot.caption = element_text(hjust = 0.5, face = "italic", size = 10),
        plot.subtitle = element_text(hjust = 0.5))
```

```{r}
#need to look carefully at this to make sure that credit is being properly assigned
#where a takeaway occurs, xAVzw == 0, make that a negative action
#where a goal is taken, should be xAVxy (quality of shot)
# zone_score <- results %>%
#   dplyr::select(id, xAV) %>%
#   rename(c("xAVxy" = "xAV"))
# 
# data <- right_join(data, zone_score, by = c("id"))
# 
# zone_score2 <- results %>%
#   dplyr::select(id, xAV) %>%
#   rename(c("xAVzw" = "xAV"))
# 
# use_data <- data %>%
#   filter(Event %in% c("Goal", "Shot", "Play", "Incomplete Play") & ! is.na(zone_next))
# 
# use_data <- right_join(data, zone_score2, by = c("zone_next" = "id"))
```

```{r}
zone_score <- results %>%
  dplyr::select(id, xAV) %>%
  rename(c("xAVxy" = "xAV"))
#the value of each zone

data <- right_join(data, zone_score, by = c("id"))
#now, every action has a specific value
#we'll again filter out actions that we're interested in where stuff happens
#penatlies, dump in/outs, puck recovery, zone entries are not things we're interested in
#we don't take takeaways either b/c they're recorded as their own event

data_score <- data %>%
  filter(Event %in% c("Goal", "Incomplete Play", "Play", "Shot"))

#incomplete plays are unsuccessful
#so are plays that result in a "Turnover" according to our change_poss variable
#for these, we want 0 as the next value so those passes register w/ negative value

zone_score2 <- results %>%
  dplyr::select(id, xAV) %>%
  rename(c("xAVzw" = "xAV",
           "next_zone" = "id"))

# test <- data_score %>%
#   filter(is.na(next_zone)) %>%
#   ungroup() %>%
#   count(Event)

data_score <- merge(data_score, zone_score2, by = "next_zone", all = TRUE, incomparables = 0)
zone_scorena <- results %>%
  dplyr::select(id, xAV) %>%
  rename(c("xAVzw_na" = "xAV",
           "id_two" = "id"))

data_score <- merge(data_score, zone_scorena, by = "id_two", all = TRUE, incomparables = 0)

data_score <- data_score %>%
  ungroup() %>%
  mutate(xAVzw = ifelse(Event %in% c("Shot", "Goal"), NA, 
                        ifelse(change_poss == "Turnover" | Event == "Incomplete Play", 0, 
                               ifelse(is.na(xAVzw) & (Event == "Play" & change_poss != "Turnover"), xAVzw_na,
                                      xAVzw))), 
         xAV = ifelse(Event %in% c("Shot", "Goal"), xAVxy, xAVzw - xAVxy)) %>%
  dplyr::select(game_date, Team:League, change_poss, id, next_zone, id_two, zone_next, xAVxy:xAV)
```

```{r}
test <- data_score %>%
  filter(is.na(xAV)) #should be 0 rows now
nrow(test) == 0

leaders <- data_score %>%
  group_by(League, Team, Player) %>%
  summarise(xAV = sum(xAV), 
            Events = n(),
            xAV_100 = (xAV / Events)*100) 

rlead <- data %>%
  group_by(League, Team, Player) %>%
  count(Event) %>%
  spread(Event, n, fill = 0) %>%
  mutate(`Shot%` = round(((Shot + Goal) / (Shot + Goal + Play + `Incomplete Play` + `Dump In/Out`))*100)) %>%
  #look at frequency of shots
  dplyr::select(League:Player, Shot, Goal, `Shot%`)

all_lead <- merge(leaders, rlead, all = TRUE)

xglead <- test_df %>%
  group_by(League, Team, Player) %>%
  summarise(xG = sum(xG, na.rm = TRUE))

all_lead <- merge(all_lead, xglead, all = TRUE)

all_lead <- all_lead %>%
  mutate(xg_100 = (xG / (Shot + Goal))*100)

bdc <- all_lead %>%
  filter(! is.na(xAV) & ! is.na(xG))

cor_res <- round(cor(bdc$xAV, bdc$xG), digits = 3)

write.csv(all_lead, file = "leaders.csv", row.names = FALSE)

all_lead %>%
  ggplot() +
  geom_point(aes(x = xAV, y = xG, color = `Shot%`)) +
  scale_color_gradient(low = "white", high = "purple") +
  #geom_text_repel(data = subset(all_lead, xAV >= 1.5 | xG >= 3), 
   #               aes(x = xAV, y = xG, label = Player, color = League)) +
  geom_smooth(aes(x = xAV, y = xG)) +
  #geom_abline(slope = 1.75)
  theme_minimal() +
  labs(title = "xG vs xAV",
       subtitle = paste0("Correlation: ", cor_res)) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
        plot.caption = element_text(hjust = 0.5, face = "italic", size = 10),
        plot.subtitle = element_text(hjust = 0.5),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(color = "lightgrey", fill = "lightgrey"))
```

### Mixture Model for pass clustering

```{r}
passes <- data %>%
  filter(Event %in% c("Play", "Incomplete Play")) %>%
  #ignores dump in/outs  b/c tgey are a concession of possession
  mutate(X.Coordinate = ifelse(X.Coordinate == X.Coordinate.2, X.Coordinate + 0.001, X.Coordinate),
         Direction = (Y.Coordinate.2 - Y.Coordinate) / (X.Coordinate.2 - X.Coordinate),
         Length = sqrt(((X.Coordinate.2 - X.Coordinate) ^ 2) + ((Y.Coordinate.2 - Y.Coordinate) ^ 2))) #this way we don't have NAs

direct_passes <- passes %>%
  ungroup() %>%
  filter(Detail.1 == "Direct") %>%
  mutate(Pass_type = ifelse(Detail.1 == "Direct", 0, 1),
         Pass_type = as.factor(Pass_type)) %>%
  dplyr::select(X.Coordinate, X.Coordinate.2, Y.Coordinate, Y.Coordinate.2,
                Direction, Length)

indirect_passes <- passes %>%
  ungroup() %>%
  filter(Detail.1 == "Indirect") %>%
  mutate(Pass_type = ifelse(Detail.1 == "Direct", 0, 1),
         Pass_type = as.factor(Pass_type)) %>%
  dplyr::select(X.Coordinate, X.Coordinate.2, Y.Coordinate, Y.Coordinate.2,
                Direction, Length)
```

```{r}
#####Gaussian for Direct Passes
set.seed(363)
direct_mm <- Mclust(direct_passes)
summary(direct_mm) #still 9 passes!
summary(direct_mm$BIC)

# get cluster centers
direct_centers <- direct_mm$parameters$mean %>%
  t() %>%
  as.data.frame()

# name clusters
direct_centers$Cluster <- c('Cluster 1', 'Cluster 2', 'Cluster 3',
                       'Cluster 4', 'Cluster 5', "Cluster 6",
                       "Cluster 7", "Cluster 8", "Cluster 9") 

direct_preds <- as.data.frame(predict(direct_mm, direct_passes))
direct_augment <- augment(direct_mm, direct_passes)

direct_augment %>%
  rink() +
  geom_segment(aes(x = X.Coordinate, xend = X.Coordinate.2,
                   y = Y.Coordinate, yend = Y.Coordinate.2, color = .class)) +
  geom_point(aes(x = X.Coordinate, y = Y.Coordinate)) + #using X.Coordinate is where the pass was released from
  facet_wrap(~.class) +
  theme_minimal()
```
```{r}
###########Gaussian for Indirect Passes
set.seed(365)
indirect_mm <- Mclust(indirect_passes)
summary(indirect_mm) #down to 8 passes

# get cluster centers
indirect_centers <- indirect_mm$parameters$mean %>%
  t() %>%
  as.data.frame()

# name clusters
indirect_centers$Cluster <- c('Cluster 1', 'Cluster 2', 'Cluster 3',
                       'Cluster 4', 'Cluster 5', "Cluster 6", "Cluster 7", "Cluster 8") 

indirect_preds <- as.data.frame(predict(indirect_mm, indirect_passes))
indirect_augment <- augment(indirect_mm, indirect_passes)

indirect_passes2 <- passes %>%
  ungroup() %>%
  filter(Detail.1 == "Indirect") %>%
  mutate(Pass_type = ifelse(Detail.1 == "Direct", 0, 1),
         Pass_type = as.factor(Pass_type))

indirect_passes2$Cluster <- indirect_augment$.class


direct_passes2 <- passes %>%
  ungroup() %>%
  filter(Detail.1 == "Direct") %>%
  mutate(Pass_type = ifelse(Detail.1 == "Direct", 0, 1),
         Pass_type = as.factor(Pass_type))

direct_passes2$Cluster <- direct_augment$.class

pass_w_cluster <- rbind(direct_passes2, indirect_passes2) %>%
  mutate(xG = NA)

non_passes <- data %>%
  filter(! Event %in% c("Play", "Incomplete Play", "Shot", "Goal")) %>%
  #ignores dump in/outs  b/c tgey are a concession of possession
  mutate(X.Coordinate = NA,
         Direction = NA,
         Length = NA,
         Cluster = NA, 
         Pass_type = NA,
         xG = NA)

all_data_w_cluster <- rbind(non_passes, pass_w_cluster)

all_data_w_cluster <- all_data_w_cluster %>%
  arrange(desc(Play_event))

df <- as.data.frame(all_data_w_cluster)

write.csv(non_passes, file = "non_passes_shots.csv", row.names = FALSE)
write.csv(pass_w_cluster, file = "pass_cluster.csv", row.names = FALSE)

shots <- data %>%
  filter(Event %in% c("Shot", "Goal")) %>%
  #ignores dump in/outs  b/c tgey are a concession of possession
  mutate(X.Coordinate = NA,
         Direction = NA,
         Length = NA,
         Cluster = NA, 
         Pass_type = NA) %>%
  arrange(desc(row_num))

xg_shots <- test_df %>%
  filter(Event %in% c("Shot", "Goal")) %>%
  arrange(desc(row_num))

shots$xG <- xg_shots$xG

write.csv(shots, file = "shots.csv", row.names = FALSE)

indirect_augment %>%
  rink() +
  geom_segment(aes(x = X.Coordinate, xend = X.Coordinate.2,
                   y = Y.Coordinate, yend = Y.Coordinate.2, color = .class)) +
  geom_point(aes(x = X.Coordinate, y = Y.Coordinate)) + #using X.Coordinate is where the pass was released from
  facet_wrap(~.class)
```

